---
title: "Hazard of MI or Stroke After UTI With and Without Positive Microbiology"
output: html_document
---

# Introduction

This document presents results of the analysis of acute cardiovascular events (MI and stroke) with GP diagnosed (suspected) UTI as exposure. Suspected UTIs with positive culture are compared to suspected UTIs with no microbiological evidence of UTI.
Number of events are all rounded to the nearest ten.

# Data

Data Tables: SAILW0972V.VB_WLGP_SUB

Derived from:  

SAIL0972V.WRRS_OBSERVATION_RESULT_20211019  
SAIL0972V.WRRS_OBSERVATION_REQUEST_20211019  
SAIL0972V.ADDE_DEATHS_20210428  
SAIL0972V.WDSD_CLEAN_ADD_GEOG_CHAR_LSOA2011_20210502  
SAIL0972V.WDSD_AR_PERS_20210502  
SAIL0972V.WDSD_CLEAN_ADD_WALES_20210502  
SAIL0972V.PEDW_SPELL_20211101  
SAIL0972V.PEDW_EPISODE_20211101  
SAIL0972V.PEDW_DIAG_20211101  
SAIL0972V.WLGP_GP_EVENT_CLEANSED_20220301  
SAIL0972V.WLGP_CLEAN_GP_REG_MEDIAN_20220301


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
#required packages
library(RODBC)
library(tcltk)
library(xlsx)
library(dplyr)
library(lubridate)
library(kableExtra)
library(survival)
library(coxme)
library(mice)
library(Publish)
library(lava)
library(ggplot2)
theme_set(theme_bw())#sets theme for plots - white background
```

```{r Connect}
# connect to SAIL SQL
getLogin<-function(userName=''){  
  require(tcltk);  
  wnd<-tktoplevel();
  tclVar(userName)->user;
  tclVar("")->passVar;
  #Label  
  
  #Username box  
  tkgrid(tklabel(wnd,text="Username:"));  
  tkgrid(tkentry(wnd,textvariable=user)->passBox);  
  
  #Password box  
  tkgrid(tklabel(wnd,text="Password:"));  
  tkgrid(tkentry(wnd,textvariable=passVar,show="*")->passBox);  
  #Hitting return will also submit password  
  tkbind(passBox,"<Return>",function() tkdestroy(wnd));  
  #OK button  
  tkgrid(tkbutton(wnd,text="OK",command=function() tkdestroy(wnd)));  
  #Wait for user to click OK  
  tkwait.window(wnd);  
  password<-tclvalue(passVar);
  userName<-tclvalue(user);
  return(c(userName,password));  
} 

login <- getLogin('<username>') # change user name and provide SAIL password in pop up box 
channel <- odbcConnect('PR_SAIL',login[1],login[2]) # connect to DB2/SAIL 
rm(login) # hides password 
```

```{r ReadData}

Cohort<- sqlQuery(channel, "SELECT * FROM SAILW0972V.VB_WLGP_SUB")
Cohort$UTI_DT<-Cohort$DIAG_DT
Cohort$MI_WITHIN90[is.na(Cohort$MI_WITHIN90)]<-0
Cohort$STROKE_WITHIN90[is.na(Cohort$STROKE_WITHIN90)]<-0

Cohort[c("UTI_OUTCOME")]<-lapply(Cohort[c("UTI_OUTCOME")],as.factor)
Cohort$UTI_OUTCOME<-relevel(Cohort$UTI_OUTCOME,ref="No microbiological evidence of UTI")

Cohort$WIMD<-as.numeric(substr(Cohort$WIMD_2019_QUINTILE_DESC,1,1))

```


```{r Timevars}
#create time variables
#calculate follow up time - time from UTI to earliest of MI/stroke, end, or 90 days

#time from  UTI to MI
Cohort$MITime<-Cohort$MI_DT-Cohort$UTI_DT
#if end date within 90 days of UTI, and no MI, use time to end
Cohort$MITime[((Cohort$INC_END_DT-Cohort$UTI_DT)<90)&is.na(Cohort$MITime)]<-Cohort$INC_END_DT[((Cohort$INC_END_DT-Cohort$UTI_DT)<90)&is.na(Cohort$MITime)]-Cohort$UTI_DT[((Cohort$INC_END_DT-Cohort$UTI_DT)<90)&is.na(Cohort$MITime)]
#if no MI, use 90 days
Cohort$MITime[is.na(Cohort$MITime)]<-90
#if MI after 90 days, use 90 days
Cohort$MITime[Cohort$MITime>90]<-90

Cohort$StrokeTime<-Cohort$STROKE_DT-Cohort$UTI_DT
Cohort$StrokeTime[((Cohort$INC_END_DT-Cohort$UTI_DT)<90)&is.na(Cohort$StrokeTime)]<-Cohort$INC_END_DT[((Cohort$INC_END_DT-Cohort$UTI_DT)<90)&is.na(Cohort$StrokeTime)]-Cohort$UTI_DT[((Cohort$INC_END_DT-Cohort$UTI_DT)<90)&is.na(Cohort$StrokeTime)]
Cohort$StrokeTime[is.na(Cohort$StrokeTime)]<-90
Cohort$StrokeTime[Cohort$StrokeTime>90]<-90

```


```{r CompleteCaseIPTW}
#complete case
#propensity scores +IPTW
Cohortcc<-Cohort[!is.na(Cohort$BMI_VAL)&!is.na(Cohort$ETHNIC)&!is.na(Cohort$WIMD_2019_QUINTILE_DESC),]

#propensity scores
Cohortcc$PropenScores<-glm(UTI_OUTCOME~AGE_AT_DIAG+BMI_VAL+DEMENTIA+WIMD_2019_QUINTILE_DESC+DIABETES+ETHNIC+EFI+HRT+RENAL_DISEASE+PARKINSONS+MND+MS+GNDR_CD+PRIM_IMMDEF+AUTOIMM, family=binomial(link="logit"), data=Cohortcc)$fitted
#IPTW
Cohortcc$weight<-ifelse(Cohortcc$UTI_OUTCOME=="Confirmed UTI",1/(Cohortcc$PropenScores),1/(1-Cohortcc$PropenScores))

```

```{r CompleteCase,include=FALSE}
#complete case
#Cox prop hazards
#90 days

ccAnalysis<-function(Cohortcc){

#set up results table
rescc<-matrix(nrow=2,ncol=4)
rescc<-data.frame(rescc)
colnames(rescc)<-c("HR","Crude","NEventsConfirmed","NEventsNoMicroEv")
rownames(rescc)<-c("MI","Stroke")


MIHR <- coxme(Surv(MITime, MI_WITHIN90) ~ UTI_OUTCOME + (1|ALF_PE), weights=weight, data = Cohortcc)
StrokeHR<- coxme(Surv(StrokeTime, STROKE_WITHIN90) ~ UTI_OUTCOME + (1|ALF_PE), weights=weight, data = Cohortcc)

MICR <- coxph(Surv(MITime, MI_WITHIN90) ~ UTI_OUTCOME, data = Cohortcc)
StrokeCR<- coxph(Surv(StrokeTime, STROKE_WITHIN90) ~ UTI_OUTCOME, data = Cohortcc)


rescc[1,1]<-paste(round(exp(MIHR$coefficients[1]),2)," [",round(exp(confint(MIHR)[1,1]),2),",",round(exp(confint(MIHR)[1,2]),2), "]" ,sep="")
rescc[1,2]<-paste(round(exp(MICR$coefficients[1]),2)," [",round(exp(confint(MICR)[1,1]),2),",",round(exp(confint(MICR)[1,2]),2), "]" ,sep="")
rescc[1,3]<-round(sum(Cohortcc$MI_WITHIN90==1&Cohortcc$UTI_OUTCOME=="Confirmed UTI"),-1)
rescc[1,4]<-round(sum(Cohortcc$MI_WITHIN90==1&Cohortcc$UTI_OUTCOME=="No microbiological evidence of UTI"),-1)

rescc[2,1]<-paste(round(exp(StrokeHR$coefficients[1]),2)," [",round(exp(confint(StrokeHR)[1,1]),2),",",round(exp(confint(StrokeHR)[1,2]),2), "]" ,sep="")
rescc[2,2]<-paste(round(exp(StrokeCR$coefficients[1]),2)," [",round(exp(confint(StrokeCR)[1,1]),2),",",round(exp(confint(StrokeCR)[1,2]),2), "]" ,sep="")
rescc[2,3]<-round(sum(Cohortcc$STROKE_WITHIN90==1&Cohortcc$UTI_OUTCOME=="Confirmed UTI"),-1)
rescc[2,4]<-round(sum(Cohortcc$STROKE_WITHIN90==1&Cohortcc$UTI_OUTCOME=="No microbiological evidence of UTI"),-1)


return(rescc)
}


```

```{r GetPooledFunc}
#full data (imputed)

#get results (where mHR is object produced by running cox on multiply imputed datasets, that have previously has propensity score and weights calculated and added) by:
#res<-as.mira(mHR)
#pool(res)# doesn't work with coxme

#tabulate parameter estimates and their variances - 
#se given in model results is calculated by print function, so can't be extracted. So calculate using the code within the print function. 

#generalised for any number of coefficients
getPooledHR<-function(mHR){
  nvar<-length(mHR$`1`$coefficients)
  nfrail<-nrow(mHR$`1`$var)-nvar
  
  res<-matrix(c(
  mHR$`1`$coefficients, diag(mHR$`1`$var)[nfrail+1:nvar],
mHR$`2`$coefficients,  diag(mHR$`2`$var)[nfrail+1:nvar],
mHR$`3`$coefficients,  diag(mHR$`3`$var)[nfrail+1:nvar],
mHR$`4`$coefficients,  diag(mHR$`4`$var)[nfrail+1:nvar],
mHR$`5`$coefficients,  diag(mHR$`5`$var)[nfrail+1:nvar],
mHR$`6`$coefficients, diag(mHR$`6`$var)[nfrail+1:nvar],
mHR$`7`$coefficients,  diag(mHR$`7`$var)[nfrail+1:nvar],
mHR$`8`$coefficients,  diag(mHR$`8`$var)[nfrail+1:nvar],
mHR$`9`$coefficients,  diag(mHR$`9`$var)[nfrail+1:nvar],
mHR$`10`$coefficients,  diag(mHR$`10`$var)[nfrail+1:nvar]
  ),byrow=T,ncol=2*nvar)

  resPool<-matrix(nrow=2,ncol=nvar)
  for (i in 1:nvar)
  {
    resPool[1,i]<-pool.scalar(res[,i],res[,nvar+i],n=mHR$`1`$n[2])$qbar
    resPool[2,i]<-pool.scalar(res[,i],res[,nvar+i],n=mHR$`1`$n[2])$ubar
  }

  #qbar = pooled estimates, ubar=mean of variances
  coef<-resPool[1,]#log(HR)
  se<-sqrt(resPool[2,])
  ciu<-coef+1.96*se
  cil<-coef-1.96*se
  #return(c(exp(coef),exp(cil),exp(ciu)))
  return(paste(round(exp(coef),2), " [", round(exp(cil),2), ",", round(exp(ciu),2), "]", sep=""))
}

```

```{r ImputationIPTW}
#multiple imputation of missing data 
#only include required columns - for propensity score, times and outcomes
CohortIm<-Cohort[,c("MITime","StrokeTime","MI_WITHIN90","STROKE_WITHIN90","UTI_OUTCOME", "AGE_AT_DIAG", "BMI_VAL", "DEMENTIA", "WIMD", "DIABETES", "ETHNIC", "EFI", "HRT", "RENAL_DISEASE", "PARKINSONS", "MND", "MS", "GNDR_CD","PRIM_IMMDEF","AUTOIMM")]
miCohort<-mice(CohortIm,m=10, print=FALSE)

miCohortLong<-complete(miCohort, "long", include=TRUE)#imputed datasets in long format, including original dataset, needed to turn back into mids object

#Propensity score model, fit to multiply imputed datasets
IPTW<-by(miCohortLong, miCohortLong$.imp, function(x) {
    PropScore<-with(x, glm(UTI_OUTCOME~AGE_AT_DIAG+BMI_VAL+DEMENTIA+WIMD+DIABETES+ETHNIC+EFI+HRT+RENAL_DISEASE+PARKINSONS+MND+MS+GNDR_CD+PRIM_IMMDEF+AUTOIMM, family=binomial(link="logit"))$fitted)
    
    #calculate weight
    weight<-ifelse(x$UTI_OUTCOME=="Confirmed UTI",1/(PropScore),1/(1-PropScore))
    
    return(weight)
})

IPTW$`0`<-rep(NA,length(IPTW$`0`))

miCohortLong$weight<-unlist(IPTW)
miCohortLong$PID<-rep(Cohort$ALF_PE,11)#add ID back in
miCohort<-as.mids(miCohortLong)

#per Rubin's rule, the point estimate of a prediction is the average of the m estimates (paper: obtaining predictions from models fit to multiply imputed data, Miles 2015)



```

```{r MainAnalysis,include=FALSE}
#Cox prop hazards
#90 days

Analysis<-function(Cohort){

#set up results table
results<-matrix(nrow=2,ncol=4)
results<-data.frame(results)
colnames(results)<-c("HR","Crude","NEventsConfirmed","NEventsNoMicroEv")
rownames(results)<-c("MI","Stroke")

miCohortLong<-complete(miCohort, "long")

MIHR<-by(miCohortLong, miCohortLong$.imp, function(x) { #apply function to miRTIlongw, split by factor .imp
  #hazard ratios
  PropHazard <- with(x,coxme(Surv(MITime, MI_WITHIN90) ~ UTI_OUTCOME + (1|PID),weights=weight))
  return(PropHazard)
})

MItest<-cox.zph(MIHR$`1`)

StrokeHR<-by(miCohortLong, miCohortLong$.imp, function(x) { #apply function to miRTIlongw, split by factor .imp
  #hazard ratios
  PropHazard <- with(x,coxme(Surv(StrokeTime, STROKE_WITHIN90) ~ UTI_OUTCOME + (1|PID),weights=weight))
  return(PropHazard)
})

Stroketest<-cox.zph(StrokeHR$`1`)

MICR <- with(Cohort, coxph(Surv(MITime, MI_WITHIN90) ~ UTI_OUTCOME))
StrokeCR<- with(Cohort, coxph(Surv(StrokeTime, STROKE_WITHIN90) ~ UTI_OUTCOME))

results[1,1]<-getPooledHR(MIHR)

res<-summary(pool(MICR),conf.int=T,exponentiate = T)
results[1,2]<-paste(round(res$estimate[1],2)," [",round(res$`2.5 %`[1] ,2),",",round(res$`97.5 %`[1],2), "]" ,sep="")


results[1,3]<-round(sum(Cohort$data$MI_WITHIN90==1&Cohort$data$UTI_OUTCOME=="Confirmed UTI"),-1)

results[1,4]<-round(sum(Cohort$data$MI_WITHIN90==1&Cohort$data$UTI_OUTCOME=="No microbiological evidence of UTI"),-1)

results[2,1]<-getPooledHR(StrokeHR)

res<-summary(pool(StrokeCR),conf.int=T,exponentiate = T)
results[2,2]<-paste(round(res$estimate[1],2)," [",round(res$`2.5 %`[1] ,2),",",round(res$`97.5 %`[1],2), "]" ,sep="")

results[2,3]<-round(sum(Cohort$data$STROKE_WITHIN90==1&Cohort$data$UTI_OUTCOME=="Confirmed UTI"),-1)
results[2,4]<-round(sum(Cohort$data$STROKE_WITHIN90==1&Cohort$data$UTI_OUTCOME=="No microbiological evidence of UTI"),-1)


return(list(results,MItest,Stroketest))
}


```

## Cox Model
coxme(Surv(TimeToEvent, EventStatus) ~ UTI_outcome + (1|ALF_PE), weights=weight)

## Hazard Ratios
Includes all data, with missing values in WIMD Quintile, BMI and ethnicity imputed. Propensity scores and IPTW are calculated on all ten datasets, then the Cox model is fitted and results are pooled. Hazard ratios are the hazard for those with confirmed UTI, compared to those with no microbiological evidence of UTI. Hazard ratios are given for 0-90 days, along with the corresponding confidence intervals. Follow up time is total follow up time in years.

```{r MainAnalysisRes}

#followup time
Follow<-matrix(nrow=2,ncol=2)
Follow<-data.frame(Follow)
colnames(Follow)<-c("Confirmed","No micro ev")
rownames(Follow)<-c("MI","Stroke")

Follow[1,1]<-round(as.numeric(sum(Cohort$MITime[Cohort$UTI_OUTCOME=="Confirmed UTI"]))/365,2)
Follow[1,2]<-round(as.numeric(sum(Cohort$MITime[Cohort$UTI_OUTCOME=="No microbiological evidence of UTI"]))/365,2)
Follow[2,1]<-round(as.numeric(sum(Cohort$StrokeTime[Cohort$UTI_OUTCOME=="Confirmed UTI"]))/365,2)
Follow[2,2]<-round(as.numeric(sum(Cohort$StrokeTime[Cohort$UTI_OUTCOME=="No microbiological evidence of UTI"]))/365,2)

kable_styling(kable_input = kbl(Follow,caption = "Follow-up time in years"),full_width = FALSE,position = "left", font_size = 12 )

full<-Analysis(miCohort)

#full data results
kable_styling(kable_input = kbl(full[[1]],caption = ""),full_width = FALSE,position = "left", font_size = 12 )

```


## Proportional Hazards Assumption 

The proportional hazards assumption was assessed visually using Kaplan Meier plots (not shown), and by a formal score test for slope = 0 (Grambsch, P. and Therneau, T. 1994).

MI
```{r }
full[[2]]
```

Stroke
```{r }
full[[3]]
```

```{r prophaz}

#Kaplan Meier plots 

#plot(survfit(Surv(miCohortLong$MITime, miCohortLong$MI_WITHIN90) ~ miCohortLong$UTI_OUTCOME, weights=miCohortLong$weight), ylim=c(0.99,1),main="MI",col=c(1,2),lty=c(1,2))



#plot(survfit(Surv(miCohortLong$StrokeTime, miCohortLong$STROKE_WITHIN90) ~ miCohortLong$UTI_OUTCOME, weights=miCohortLong$weight), ylim=c(0.99,1),main="Stroke",col=c(1,2),lty=c(1,2))



```

## Complete Case
Uses only individuals with no missing data, and therefore no imputation. Analysis is otherwise as previously.

```{r CompleteCaseRes}
#followup time
FollowCC<-matrix(nrow=2,ncol=2)
FollowCC<-data.frame(FollowCC)
colnames(FollowCC)<-c("Confirmed","No micro ev")
rownames(FollowCC)<-c("MI","Stroke")

FollowCC[1,1]<-round(as.numeric(sum(Cohortcc$MITime[Cohortcc$UTI_OUTCOME=="Confirmed UTI"]))/365,2)
FollowCC[1,2]<-round(as.numeric(sum(Cohortcc$MITime[Cohortcc$UTI_OUTCOME=="No microbiological evidence of UTI"]))/365,2)
FollowCC[2,1]<-round(as.numeric(sum(Cohortcc$StrokeTime[Cohortcc$UTI_OUTCOME=="Confirmed UTI"]))/365,2)
FollowCC[2,2]<-round(as.numeric(sum(Cohortcc$StrokeTime[Cohortcc$UTI_OUTCOME=="No microbiological evidence of UTI"]))/365,2)

kable_styling(kable_input = kbl(FollowCC,caption = "Follow-up time in years"),full_width = FALSE,position = "left", font_size = 12 )

#complete case results
rescc<-ccAnalysis(Cohortcc)

kable_styling(kable_input = kbl(rescc,caption = ""),full_width = FALSE,position = "left", font_size = 12 )
```
